Sparse spectral and $p$-finite element methods on disk-slices and rhombi I: zero boundary conditions

### 2019-03-07 papers/halfdisk/halfdisk.tex
bmatrix -> pmatrix

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: \sotodo{Why bmatrix, not pmatrix?}\bstodo{No reason, I am not familiar with pmatrix}
\begin{align}
\bigPab_n := \begin{bmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{bmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\bigPab_2 \\
		\vdots \\
	\end{bmatrix}
\end{align}
\sotodo{Why not just make $\bigPab$ infinite for now?}and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bigPab(x,y) = x \: \bigPab(x,y), \quad J_y^{(a,b)} \: \bigPab(x,y) = y \: \bigPab(x,y).
\end{align}
Then $J_x^{(a,b)}, J_y^{(a,b)}$ have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{bmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{bmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{bmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{bmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^T \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{bmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{bmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{bmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{bmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align}

@@

Combining each system in (71) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{bmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{bmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each $n = 0,1,2\dots$:
\begin{align}
A_n &:= \begin{bmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{bmatrix}
		C^x_n \\
		C^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{bmatrix}
		B^x_n \\
		B^y_n
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{bmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{bmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

@@

Note that we can define an explicit \(\Dnt\) as follows. For \(n\) even:
\begin{align}
\Dnt := \begin{bmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & & &  \\
		& \ddots & & & \\
		& & \frac{1}{\alphaab_{n+1,n,1}} & \\
		0 & \hdots & 0 & \eta_{m-1} & \hdots & 0 & \eta_1 & 0 & \eta_0
	    \end{bmatrix},
\end{align}
where
\begin{align}
m &= \frac{n}{2} + 1, \quad \eta_0 = \frac{1}{\betaab_{n,n,6}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-1.
\end{align}

@@

\begin{align}
\Dnt := \begin{bmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & &  \\
		& \ddots & & &  \\
		& & \ddots & & \\
		& & & \frac{1}{\alphaab_{n+1,n,1}} & \\
		\xi & 0 & \hdots & 0 & 0 & \eta_{m-2} & \hdots & 0 & \eta_0
	    \end{bmatrix},
\end{align}
 where
\begin{align}
m &= \frac{n+1}{2} + 1,\quad  \eta_0 = \frac{1}{\betaab_{n,n,6}}, \quad \xi = -\frac{\betaab_{n,1,5} \: \eta_{m-2}}{\alphaab_{n+1,0,1}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-2.
\end{align}
It follows that we can apply $D_n^T$ in $O(n)$ complexity, and thereby construct $\bigPab_{n}(x,y)$ in optimal $O(n^2)$ complexity.

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) of degree $N$ we can express \(f\) by
\begin{align*}
f(x,y) = \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^T \: \mathbf{f}_n
\end{align*}
where
\begin{align*}
\bigP_n(x,y) &:= \begin{bmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align*}
and where
\begin{align*}
\mathbf{f}_n &:= \begin{bmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \quad
f_{n,k} := \frac{\langle f, \: \Pnk^{(a,b)} \rangle_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align*}

@@

Define $\Lambda^{(a,b)} := \langle \bigP^{(a,b)}, \: {\bigP^{(a,b)}}^T \rangle_{W^{(a,b)}}$ where $W^{(a,b)}$ is the weight with which the OPs in $\bigP^{(a,b)}$ are orthogonal with respect to. Note that due to orthogonality this is a diagonal matrix. We can choose a basis for $V_h$ by using the weighted orthogonal polynomials on $\element$ with parameters $a = b = 1$:
\begin{align}
\bigWii(x,y) &:= \begin{bmatrix}
		\bigWii_0(x,y) \\
		\bigWii_1(x,y) \\
		\bigWii_2(x,y) \\
		\vdots \\
		\bigWii_p(x,y)
	\end{bmatrix}, \\
\bigWii_n(x,y) &:= \begin{bmatrix}
		\Wii(x,y) \: \Pii_{n,0}(x,y) \\
		\vdots \\
		\Wii(a,y) \: \Pii_{n,n}(x,y)
	\end{bmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,p,
\end{align}
and rewrite (73) in matrix form:
\begin{align}
	a(u_h,v_h) &= \int_\element \nabla u_h \cdot \nabla v_h \: d\mathbf{x} \\
	&= \int_\element \begin{bmatrix}
					\partial_x v_h \\
					\partial_y v_h
				\end{bmatrix}^T 
				\begin{bmatrix}
					\partial_x u_h \\
					\partial_y u_h
				\end{bmatrix}
				\\
	&= \int_\element \begin{bmatrix}
					\bigPoo^T \Wii_x \mathbf{v} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{v}
				\end{bmatrix}^T 
				\begin{bmatrix}
					\bigPoo^T \Wii_x \mathbf{u} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}
				\end{bmatrix}
				\\
	&= \int_\element \Big( \mathbf{v}^T {\Wii_x}^T \bigPoo \bigPoo^T \Wii_x \mathbf{u} \nonumber \\
					& \quad \quad \quad + \mathbf{v}^T ({T^{(1,0)\to(0,0)} \Wii_y})^T \bigPoo \bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}  \Big) \: d\mathbf{x} \\
	&= \mathbf{v}^T \: \Big( {\Wii_x}^T \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^T \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y \Big) \: \mathbf{u}
\end{align}
where $\mathbf{u}, \mathbf{v}$ are the coefficient vectors of the expansions of $u_h, v_h \in V_h$ respectively in the $V_h$ basis ($\bigWii$ OPs), and
\begin{align}
	L(v_h) &= \int_\element \: v_h \: f \: d\mathbf{x} \\
	&= \int_\element \: \mathbf{v}^T \: \bigWii \: \bigPii^T \: \mathbf{f} \: d\mathbf{x} \\
	&= \mathbf{v}^T \: \langle \bigPii, {\bigPii}^T \rangle_{\Wii} \: d\mathbf{x} \\
	&= \mathbf{v}^T \Lambda^{(1,1)} \: \mathbf{f},
\end{align}
where $\mathbf{f}$ is the coefficient vector for the expansion of the function $f(x,y)$ in the $\bigPii$ OP basis.

@@

\subsection{Two-element p-FEM on the disk (without BCs)}
We now consider the case of using two elements to discretise a unit disk domain $\Omega$, that is the each element $\element$ is a half-disk, with common edge defined by the line $x=0$ for $y \in [-1,1]$. We take inspiration from \cite{karniadakis2013spectral, beuchler2006new}. To ensure continuity across the edge interface, we need to expand our basis functions to be used. We require basis functions that correspond to each edge (in this case there is only one edge), that is basis functions with support as the two elements connected by said edge (for more than two elements - and hence more than one edge - we would also stipulate that each edge's corresponding basis functions would vanish on any other edge). We then also require the remaining (interior) basis functions be such that they have support of one element only, and vanish on (each) edge. Using this, our basis functions for $\le N$-degree polynomials can then be written as
\begin{align}
	\bigQab_n(x,y) := \begin{bmatrix}
		P^{(a,b)}_{n,n}(x,y) \\
		x \: \bigPab_{n-1}(x,y)
	\end{bmatrix} \in \R^{n+1}, 
	\quad \quad 
	\bigQab := \begin{bmatrix}
		\bigQab_0 \\
		\bigQab_1 \\
		\vdots \\
		\bigQab_N
	\end{bmatrix}.
\end{align}
We note that for each $n = 0,\dots,N$, $P^{(a,b)}_{n,n}(x,y) = \rho(x)^n \: P^{(a,b)}_{n}(\frac{y}{\rho(x)})$ and so at $x=0$ these become just the Jacobi polynomials in $y$, thus span the edge interface. The polynomials that make up the $\bigQab$ basis can also be shown to have similar sparse relations and yield similar sparse operator matrices. We can then proceed to use a similar approach to section A.1 on a reference element.

@@

We can use the similar idea of projecting up from the 2D orthogonal polynomials on the half disk to 3D OPs on the hemisphere. Define $\rho(x) := (1-x^2)^\half$ for $x \in [0,1]$ and let the half disk be denoted by $\omega$, that is let
\begin{align}
	\omega := \{(x,y) \in \R^2 \quad | \quad 1 < x < 0, \: \rho(x) < y < \rho(x)\}.
\end{align}
Further, define the hemisphere by
\begin{align}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad (x,y) \in \omega, \: x^2 + y^2 + z^2 = 1\}.
\end{align}
We can define the 3D OP family on the hemisphere as the set of polynomials $\bigQab(x,y,z)$ where
\begin{align}
	\bigQab_n(x,y,z) := \begin{bmatrix}
		\bigPab_n(x,y) \\
		z \: \bigP^{(a,b+1)}_{n-1}(x,y)
	\end{bmatrix} \in \R^{2n+1}, 
	\quad \quad 
	\bigQab := \begin{bmatrix}
		\bigQab_0 \\
		\bigQab_1 \\
		\vdots \\
		\bigQab_N
	\end{bmatrix}.
\end{align}
The relations for multiplication by $x, y$ are then trivial. For multiplication by $z$, note that $z^2 = W^{(0,1)}(x,y)$ for $(x,y,z) \in \Omega$ and consider that (as the 2D OPs for any parameters form a 2D basis of $\omega$):
\begin{align}
	z \: \Pnkab(x,y) &= \sum_{m,j} c_{n,k,m,j}^{(a,b)\to(a,b+1)} \: z \: P^{(a,b+1)}_{m,j}(x,y), \\
	z^2 \: \Pnkab(x,y) &= \sum_{m,j} d_{n,k,m,j}^{(a,b)\to(a,b-1)} \: z \: P^{(a,b-1)}_{m,j}(x,y).
\end{align}
A simple calculation shows that the expansions are sparse, that is that most coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ and $d_{n,k,m,j}^{(a,b)\to(a,b-1)}$ are $0$, and are also given by
\begin{align}
	c_{n,k,m,j}^{(a,b)\to(a,b+1)} &= \frac{\ip< \Pnkab, \: P^{(a,b+1)}_{m,j} >_{W^{(a,b+1)}}}{\norm{P^{(a,b+1)}_{m,j}}^2_{W^{(a,b+1)}}} \\
	d_{n,k,m,j}^{(a,b)\to(a,b-1)} &= \frac{\ip< W^{(0,1)} \Pnkab, \: P^{(a,b-1)}_{m,j} >_{W^{(a,b-1)}}}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \nonumber \\
	&= \frac{\ip< \Pnkab, \: P^{(a,b-1)}_{m,j} >_{W^{(a,b)}}}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \nonumber \\
	&= \frac{\norm{\Pnkab}_\Wab^2}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \: c_{m,j,n,k}^{(a,b-1)\to(a,b)}.
\end{align}
The coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ and $d_{n,k,m,j}^{(a,b)\to(a,b-1)}$ are just the entries to the parameter tranformation operator matrices
\begin{align}
	\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b+1)} \Big)^T \: \bigP^{(a,b+1)}(x,y) \\
	W^{(0,1)}(x,y) \bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b-1)} \Big)^T \: \bigP^{(a,b-1)}(x,y)
\end{align}
respectively.

>>>>

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align}
\bigPab_n := \begin{pmatrix}
		P^{(a,b)}_{n,-n}(x,y) \\
		\vdots \\
		P^{(a,b)}_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bigPab := \begin{pmatrix}
		\bigPab_0 \\
		\bigPab_1 \\
		\bigPab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bigPab(x,y) = x \: \bigPab(x,y), \quad J_y^{(a,b)} \: \bigPab(x,y) = y \: \bigPab(x,y).
\end{align}
Then $J_x^{(a,b)}, J_y^{(a,b)}$ have the Banded-Block-Banded structures
\begin{align}
J_{x/y}^{(a,b)} &= \begin{pmatrix}
		B^{x/y}_0 & A^{x/y}_0 & & & & \\
		C^{x/y}_1 & B^{x/y}_1 & A^{x/y}_1 & & & \\
		& C^{x/y}_2 & B^{x/y}_2 & A^{x/y}_2  & & & \\
		& & C^{x/y}_3 & \ddots & \ddots & \\
		& & & \ddots & \ddots & \ddots \\
	\end{pmatrix}
\end{align}
where
\begin{align}
A^x_n &:= \begin{pmatrix}
		\alphaab_{n+1,0,1} & 0 & \hdots & 0 \\
		& \ddots & & \vdots & \\
		& & \alphaab_{n+1,n,1} & 0 \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^x_n &:= \begin{pmatrix}
		\alphaab_{n,0,2} & & \\
		& \ddots & \\
		& & \alphaab_{n,n,2} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)} \quad n = 0,1,2,\dots \\
C^x_n &:= \big( A^x_n \big)^T \in \R^{(n+1)\times n},  \quad n = 1,2,\dots \\ 
\nonumber \\
A^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,6} & & & \\
		\betaab_{n,1,5} & 0 & & & \\
		& \ddots & \ddots & \ddots & \\
		& & \betaab_{n,n,5}& 0 & \betaab_{n,n,6} \\
	    \end{pmatrix} \in \R^{(n+1)\times(n+2)}, \quad n = 0,1,2,\dots \\
B^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,4} & & \\
		\betaab_{n,1,3} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-1,4} \\
		& & \betaab_{n,n,3} & 0
	    \end{pmatrix} \in \R^{(n+1)\times(n+1)}  \quad n = 0,1,2,\dots \\
C^y_n &:= \begin{pmatrix}
		0 & \betaab_{n,0,2} & & \\
		\betaab_{n,1,1} & 0 & \ddots & \\
		& \ddots & \ddots & \betaab_{n,n-2,2} \\
		& & \ddots & 0 \\
		& & & \betaab_{n,n,1} \\
	    \end{pmatrix} \in \R^{(n+1)\times n}, \quad n = 1,2,\dots
\end{align}

@@

Combining each system in (71) we can write the block-tridiagonal system
\begin{align}
\renewcommand\arraystretch{1.3}
\mleft[
\begin{array}{cccc}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{array}
\mright]
\bigPab(x,y)
=
\begin{pmatrix}
	P^{(a,b)}_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{pmatrix}.
\end{align}
where we note \(P^{(a,b)}_{0,0}(x,y) \equiv H_0^{(a,b)} \: P_0^{(b,b)}\), and for each $n = 0,1,2\dots$:
\begin{align}
A_n &:= \begin{pmatrix}
		A^x_n \\
		A^y_n \\
		A^z_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+3)} \quad (n \ne N), \quad
C_n := \begin{pmatrix}
		C^x_n \\
		C^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{pmatrix}
		B^x_n \\
		B^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{pmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align}

@@

Note that we can define an explicit \(\Dnt\) as follows. For \(n\) even:
\begin{align}
\Dnt := \begin{pmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & & &  \\
		& \ddots & & & \\
		& & \frac{1}{\alphaab_{n+1,n,1}} & \\
		0 & \hdots & 0 & \eta_{m-1} & \hdots & 0 & \eta_1 & 0 & \eta_0
	    \end{pmatrix},
\end{align}
where
\begin{align}
m &= \frac{n}{2} + 1, \quad \eta_0 = \frac{1}{\betaab_{n,n,6}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-1.
\end{align}

@@

\begin{align}
\Dnt := \begin{pmatrix}
		\frac{1}{\alphaab_{n+1,0,1}} & &  \\
		& \ddots & & &  \\
		& & \ddots & & \\
		& & & \frac{1}{\alphaab_{n+1,n,1}} & \\
		\xi & 0 & \hdots & 0 & 0 & \eta_{m-2} & \hdots & 0 & \eta_0
	    \end{pmatrix},
\end{align}
 where
\begin{align}
m &= \frac{n+1}{2} + 1,\quad  \eta_0 = \frac{1}{\betaab_{n,n,6}}, \quad \xi = -\frac{\betaab_{n,1,5} \: \eta_{m-2}}{\alphaab_{n+1,0,1}},\qqand \nonumber \\
\eta_k &= -\frac{\betaab_{n,n-2(k-1),5} \: \eta_{k-1}}{\betaab_{n,n-2k,6}} \quad k = 1,\dots,m-2.
\end{align}
It follows that we can apply $D_n^T$ in $O(n)$ complexity, and thereby construct $\bigPab_{n}(x,y)$ in optimal $O(n^2)$ complexity.

@@

Fix \(a,b \in \R\). Then for any function \(f : \Omega \to \R\) of degree $N$ we can express \(f\) by
\begin{align*}
f(x,y) = \sum_{n=0}^N \bigP_n^{(a,b)}(x,y)^T \: \mathbf{f}_n
\end{align*}
where
\begin{align*}
\bigP_n(x,y) &:= \begin{pmatrix}
		P_{n,0}(x,y) \\
		\vdots \\
		P_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N,
\end{align*}
and where
\begin{align*}
\mathbf{f}_n &:= \begin{pmatrix}
		f_{n,0} \\
		\vdots \\
		f_{n,n}
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,N, \quad
f_{n,k} := \frac{\langle f, \: \Pnk^{(a,b)} \rangle_{W^{(a,b)}}}{|| \Pnk^{(a,b)} ||_{W^{(a,b)}}}
\end{align*}

@@

Define $\Lambda^{(a,b)} := \langle \bigP^{(a,b)}, \: {\bigP^{(a,b)}}^T \rangle_{W^{(a,b)}}$ where $W^{(a,b)}$ is the weight with which the OPs in $\bigP^{(a,b)}$ are orthogonal with respect to. Note that due to orthogonality this is a diagonal matrix. We can choose a basis for $V_h$ by using the weighted orthogonal polynomials on $\element$ with parameters $a = b = 1$:
\begin{align}
\bigWii(x,y) &:= \begin{pmatrix}
		\bigWii_0(x,y) \\
		\bigWii_1(x,y) \\
		\bigWii_2(x,y) \\
		\vdots \\
		\bigWii_p(x,y)
	\end{pmatrix}, \\
\bigWii_n(x,y) &:= \begin{pmatrix}
		\Wii(x,y) \: \Pii_{n,0}(x,y) \\
		\vdots \\
		\Wii(a,y) \: \Pii_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1} \quad \forall n = 0,1,2,\dots,p,
\end{align}
and rewrite (73) in matrix form:
\begin{align}
	a(u_h,v_h) &= \int_\element \nabla u_h \cdot \nabla v_h \: d\mathbf{x} \\
	&= \int_\element \begin{pmatrix}
					\partial_x v_h \\
					\partial_y v_h
				\end{pmatrix}^T 
				\begin{pmatrix}
					\partial_x u_h \\
					\partial_y u_h
				\end{pmatrix}
				\\
	&= \int_\element \begin{pmatrix}
					\bigPoo^T \Wii_x \mathbf{v} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{v}
				\end{pmatrix}^T 
				\begin{pmatrix}
					\bigPoo^T \Wii_x \mathbf{u} \\
					\bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}
				\end{pmatrix}
				\\
	&= \int_\element \Big( \mathbf{v}^T {\Wii_x}^T \bigPoo \bigPoo^T \Wii_x \mathbf{u} \nonumber \\
					& \quad \quad \quad + \mathbf{v}^T ({T^{(1,0)\to(0,0)} \Wii_y})^T \bigPoo \bigPoo^T T^{(1,0)\to(0,0)} \Wii_y \mathbf{u}  \Big) \: d\mathbf{x} \\
	&= \mathbf{v}^T \: \Big( {\Wii_x}^T \Lambda^{(0,0)} \Wii_x + ({T^{(1,0)\to(0,0)} \Wii_y})^T \Lambda^{(0,0)} T^{(1,0)\to(0,0)} \Wii_y \Big) \: \mathbf{u}
\end{align}
where $\mathbf{u}, \mathbf{v}$ are the coefficient vectors of the expansions of $u_h, v_h \in V_h$ respectively in the $V_h$ basis ($\bigWii$ OPs), and
\begin{align}
	L(v_h) &= \int_\element \: v_h \: f \: d\mathbf{x} \\
	&= \int_\element \: \mathbf{v}^T \: \bigWii \: \bigPii^T \: \mathbf{f} \: d\mathbf{x} \\
	&= \mathbf{v}^T \: \langle \bigPii, {\bigPii}^T \rangle_{\Wii} \: d\mathbf{x} \\
	&= \mathbf{v}^T \Lambda^{(1,1)} \: \mathbf{f},
\end{align}
where $\mathbf{f}$ is the coefficient vector for the expansion of the function $f(x,y)$ in the $\bigPii$ OP basis.

@@

\subsection{Two-element p-FEM on the disk (without BCs)}
We now consider the case of using two elements to discretise a unit disk domain $\Omega$, that is the each element $\element$ is a half-disk, with common edge defined by the line $x=0$ for $y \in [-1,1]$. We take inspiration from \cite{karniadakis2013spectral, beuchler2006new}. To ensure continuity across the edge interface, we need to expand our basis functions to be used. We require basis functions that correspond to each edge (in this case there is only one edge), that is basis functions with support as the two elements connected by said edge (for more than two elements - and hence more than one edge - we would also stipulate that each edge's corresponding basis functions would vanish on any other edge). We then also require the remaining (interior) basis functions be such that they have support of one element only, and vanish on (each) edge. Using this, our basis functions for $\le N$-degree polynomials can then be written as
\begin{align}
	\bigQab_n(x,y) := \begin{pmatrix}
		P^{(a,b)}_{n,n}(x,y) \\
		x \: \bigPab_{n-1}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
	\quad \quad 
	\bigQab := \begin{pmatrix}
		\bigQab_0 \\
		\bigQab_1 \\
		\vdots \\
		\bigQab_N
	\end{pmatrix}.
\end{align}
We note that for each $n = 0,\dots,N$, $P^{(a,b)}_{n,n}(x,y) = \rho(x)^n \: P^{(a,b)}_{n}(\frac{y}{\rho(x)})$ and so at $x=0$ these become just the Jacobi polynomials in $y$, thus span the edge interface. The polynomials that make up the $\bigQab$ basis can also be shown to have similar sparse relations and yield similar sparse operator matrices. We can then proceed to use a similar approach to section A.1 on a reference element.

@@

We can use the similar idea of projecting up from the 2D orthogonal polynomials on the half disk to 3D OPs on the hemisphere. Define $\rho(x) := (1-x^2)^\half$ for $x \in [0,1]$ and let the half disk be denoted by $\omega$, that is let
\begin{align}
	\omega := \{(x,y) \in \R^2 \quad | \quad 1 < x < 0, \: \rho(x) < y < \rho(x)\}.
\end{align}
Further, define the hemisphere by
\begin{align}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad (x,y) \in \omega, \: x^2 + y^2 + z^2 = 1\}.
\end{align}
We can define the 3D OP family on the hemisphere as the set of polynomials $\bigQab(x,y,z)$ where
\begin{align}
	\bigQab_n(x,y,z) := \begin{pmatrix}
		\bigPab_n(x,y) \\
		z \: \bigP^{(a,b+1)}_{n-1}(x,y)
	\end{pmatrix} \in \R^{2n+1}, 
	\quad \quad 
	\bigQab := \begin{pmatrix}
		\bigQab_0 \\
		\bigQab_1 \\
		\vdots \\
		\bigQab_N
	\end{pmatrix}.
\end{align}
The relations for multiplication by $x, y$ are then trivial. For multiplication by $z$, note that $z^2 = W^{(0,1)}(x,y)$ for $(x,y,z) \in \Omega$ and consider that (as the 2D OPs for any parameters form a 2D basis of $\omega$):
\begin{align}
	z \: \Pnkab(x,y) &= \sum_{m,j} c_{n,k,m,j}^{(a,b)\to(a,b+1)} \: z \: P^{(a,b+1)}_{m,j}(x,y), \\
	z^2 \: \Pnkab(x,y) &= \sum_{m,j} d_{n,k,m,j}^{(a,b)\to(a,b-1)} \: z \: P^{(a,b-1)}_{m,j}(x,y).
\end{align}
A simple calculation shows that the expansions are sparse, that is that most coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ and $d_{n,k,m,j}^{(a,b)\to(a,b-1)}$ are $0$, and are also given by
\begin{align}
	c_{n,k,m,j}^{(a,b)\to(a,b+1)} &= \frac{\ip< \Pnkab, \: P^{(a,b+1)}_{m,j} >_{W^{(a,b+1)}}}{\norm{P^{(a,b+1)}_{m,j}}^2_{W^{(a,b+1)}}} \\
	d_{n,k,m,j}^{(a,b)\to(a,b-1)} &= \frac{\ip< W^{(0,1)} \Pnkab, \: P^{(a,b-1)}_{m,j} >_{W^{(a,b-1)}}}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \nonumber \\
	&= \frac{\ip< \Pnkab, \: P^{(a,b-1)}_{m,j} >_{W^{(a,b)}}}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \nonumber \\
	&= \frac{\norm{\Pnkab}_\Wab^2}{\norm{P^{(a,b-1)}_{m,j}}^2_{W^{(a,b-1)}}} \: c_{m,j,n,k}^{(a,b-1)\to(a,b)}.
\end{align}
The coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ and $d_{n,k,m,j}^{(a,b)\to(a,b-1)}$ are just the entries to the parameter tranformation operator matrices
\begin{align}
	\bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b+1)} \Big)^T \: \bigP^{(a,b+1)}(x,y) \\
	W^{(0,1)}(x,y) \bigP^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b-1)} \Big)^T \: \bigP^{(a,b-1)}(x,y)
\end{align}
respectively.
